---
title: "A8_Walker_Charlie"
author: "Charlie Walker"
date: "2023-03-06"
output: pdf_document
---

Link to [Github Repository](https://github.com/moonflwrz/BIOL432_Assignment8)

Questions:

1.  What effect (if any) does garlic mustard have on the plant community? [A]
2.  What has a stronger effect on plant communities: the presence/absence of garlic mustard (in/out) or sampling population? [B]

```{r data_setup}
#loading libraries..
library(ggplot2)
library(ape)
library(ggtree)
library(vegan)
library(dplyr)
library(reshape2)
library(MASS)
library(colorBlindness)

theme_set(theme_bw())

#importing data.. 
floristDat<-read.csv(
  "FloristicSurvey.csv")

#..stealing IDs from 'Quadrate' 
IDs<-floristDat$Quadrate
```

### [A] ------------------

```{r A_setup}
#mutating all garlic columns into one..
AflorDat <- floristDat %>% 
  mutate(
    Alliaria_petiolata = rowSums(
      across(c('Rosettes', 'Bolting',
               'Budding', 'Bud_Flw',
               'Flower','Flw_Sil',
               'Sillique')))) %>%
  dplyr::select(1,11:45) %>%
  relocate(Alliaria_petiolata, 
           .before=Claytonia_virginiana)

```

#### A.1 - Stats

For question A, we're interesting in quantifying the effect (if any) that garlic mustard has on the plant community.

```{r A_stats}
#getting rid of quad
Af_binary<-AflorDat %>% 
  dplyr::select(-1)

#running a simple linear regression
Af_lim<-lm(Alliaria_petiolata~.,
           data=Af_binary)

#model selection.. 
BackSel<-stepAIC(Af_lim,
                direction="backward")
summary(BackSel)

#QA/QC
hist(residuals(BackSel), freq=F)
shapiro.test(residuals(BackSel))
plot(BackSel,1)

```

From our model summary, it looks like garlic mustard does indeed have an effect on the plant community- at least, the majority of species within the community. Out of the 27 species kept during model selection, 4 species are not effected by garlic mustard: *Acer saccharum,* Unknown_4, *Dryopteris marginalis*, and Grass.

We've got some outliers, but our data is normally distributed. Since the dataset is already so small, let's not touch it further.

#### A.2 - Distance Matrix

We normally wouldn't use nested loops for this conversion as they're not efficient, but this dataset is very small- only 30 observations -so the wait time shouldn't be too long (depending on your pc).

```{r A_distmat}

#converting into binary..
for (i in 1:nrow(Af_binary)){
  for(j in 1:ncol(Af_binary)){
    if(Af_binary[i,j]>0){
      Af_binary[i,j]<-1
    }
  }
}

row.names(
  Af_binary)<-IDs

#calculating pairwise distance..
Af_dist<-dist(
  Af_binary, method="binary")

#converting into matrix & melting..
Af_distmat<-Af_dist %>% 
  as.matrix() %>% 
  melt()

#plotting.. 
ggplot(data=Af_distmat, aes(
  x=Var1, y=Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradientn(
    colors=c("#785EF0", "#DC267F",
            "#FE6100", "#FFB000")) +
  theme(axis.text.x=element_text(
    angle=90, hjust=1, vjust=0.5))

```

Looking at our distance matrix, all of the 13ox quadrates are quite different from the rest. Looking at `AflorDat`, the 13ox quadrates mostly contain grass.

#### A. 3 - NMDS

```{r A_NMDS}

set.seed(3)
NMDSdat<-metaMDS(Af_dist,k=2,
                 trymax=25)

#Data for plotting.. 
Af_PDat<-data.frame(
  NMDS1=NMDSdat$points[,1],
  NMDS2=NMDSdat$points[,2],
  SampleID=row.names(Af_binary))

#plotting..
ggplot(data=Af_PDat,
       aes(colour=IDs,
           alpha=I(0.6))) +
         geom_point(x=Af_PDat$NMDS1,
                    y=Af_PDat$NMDS2) +
  xlim(-0.4, 0.7) +
  ylim(-0.3, 0.3) +
  theme(axis.text=element_blank())
```

Take note of the cluster at the bottom- those 6 quadrates are very similar to each other.

#### A.4 - Plot

```{r A_plot, fig.cap="Above displays the relationship between garlic mustard (\\textit{Alliaria petiolata}) and neighbouring plant communities found at the Queen's University Biological Station (QUBS), represented by a count of individuals. Data is coloured & organized by sampling quadrate."}

#further simplifying.. 
AfplotDat<-AflorDat %>% 
  mutate(flwr_community= rowSums(
    across(c(2:3, 6:25, 26)))) %>% 
    dplyr::select(1:2, 37)


#plotting
Aplot<-AfplotDat %>% 
  ggplot(aes(x=Alliaria_petiolata,
             y=flwr_community)) +
  geom_point(aes(colour=Quadrate))
print(Aplot)
```

It appears as if the majority of quadrates' flower communities kept within a range of 0-200 with a complementary garlic mustard range of approx. 0-80.

Quadrates that fell out of that range saw extreme gains within both their flower communites & counts of garlic mustard.

### [B] ------------------

```{r B_setup}
#Dropping garlic columns
BflorDat<- floristDat %>% 
  dplyr::select(1:3, 11:44)
  
```

#### B.1 - Stats

For question B, we're interested in determine which explanatory variable (garlic mustard presence or sampling population) has a stronger effect on plant communities.

Let's run two linear models and then compare them to determine which fits our data better.

```{r B_stats}
#Recoding Location.. 
BflorDat$Location<-recode(
  BflorDat$Location,
       o = 0,
       i = 1)

#Sampling population------
B_popDat<- BflorDat %>% 
  dplyr::select(2, 4:37)

Bp_lim<-lm(Population~.,
           data=B_popDat)

Bp_Back<-stepAIC(Bp_lim,
                direction="backward")
summary(Bp_Back)
AIC(Bp_Back)

#Garlic Presence----------
B_localDat<- BflorDat %>% 
  dplyr::select(3:37)

Bl_lim<-lm(Location~.,
           data=B_localDat)

Bl_Back<-stepAIC(Bl_lim,
                direction="backward")
summary(Bl_Back)
AIC(Bl_Back)
```

By comparing model summaries & AIC scores, it appears that sampling population has a stronger effect on plant communities.

#### B.2 - Distance Matrix

```{r B_distmat}

#same loop, differing dataset.. 
Bf_binary<-B_popDat
for (i in 1:nrow(Bf_binary)){
  for(j in 1:ncol(Bf_binary)){
    if(Bf_binary[i,j]>0){
      Bf_binary[i,j]<-1
    }
  }
}

row.names(
  Bf_binary)<-IDs

#calculating pairwise distance..
Bf_dist<-dist(
  Bf_binary, method="binary")

#converting into matrix & melting..
Bf_distmat<-Bf_dist %>% 
  as.matrix() %>% 
  melt()

#plotting.. 
ggplot(data=Bf_distmat, aes(
  x=Var1, y=Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradientn(
    colors=c("#785EF0", "#DC267F",
            "#FE6100", "#FFB000")) +
  theme(axis.text.x=element_text(
    angle=90, hjust=1, vjust=0.5))

```

It appears that in terms of our B dataset, most of the 3xx quadrates (save for 3i1) are identical. Additionally, quadrates in close proximity seem to be similar as well.

Apart from that, the majority of quadrates are considerably different (\>0.5).

#### B.3 - NMDS

```{r B_NMDS}
set.seed(3)
NMDSdat<-metaMDS(Bf_dist,k=2,
                 trymax=25)

#Data for plotting.. 
Bf_PDat<-data.frame(
  NMDS1=NMDSdat$points[,1],
  NMDS2=NMDSdat$points[,2],
  SampleID=row.names(Bf_binary))

#plotting..
ggplot(data=Bf_PDat,
       aes(colour=IDs,
           alpha=I(0.6))) +
         geom_point(x=Bf_PDat$NMDS1,
                    y=Bf_PDat$NMDS2) +
  xlim(-0.4, 0.7) +
  ylim(-0.3, 0.3) +
  theme(axis.text=element_blank())
```

Quadrates 14o2, 14o1, and 14i3 are very similar- as well as the previous similarities already outlined.

#### B.4 - Plot

```{r B_plot, fig.cap="Above displays the measure of non-garlic mustard (\\textit{Alliaria petiolata}) species (flwr_community) found at the Queen's University Biological Station (QUBS), represented as a count of individuals. Data is organized by sampling population ID, and coloured by sampling quadrate."}
#further simplifying.. 
BfplotDat<-BflorDat %>% 
  mutate(flwr_community= rowSums(
    across(c(2:27)))) %>% 
   dplyr::select(1:2, 38)
   
BfplotDat$Population<-as.factor(
  BfplotDat$Population)

#plotting
Bplot<-BfplotDat %>% 
  ggplot(aes(x=Population,
             y=flwr_community)) +
  geom_point(aes(colour=Quadrate))
print(Bplot)
```

Population 7 seems to have the highest measure of non-garlic mustard species within it's assigned quadrates. Interestingly, quadrates assigned from population 7 had both no contact (7o1, 7o2, 3o2) and high contact (7i1, 7i2, 3o3) with garlic mustard- with those in high contact with garlic mustard measured at a higher abundance.
