---
title: "A9_Walker_Charlie"
author: "Charlie Walker"
date: "2023-03-13"
output:
  pdf_document: default
  html_document: default
---

Link to [Github Repository](https://github.com/moonflwrz/BIOL432_Assignment9)

```{r data_setup}
#loading libraries..
library(doParallel)
library(parallel)
library(foreach)

#Grabbing # of cores & making cluster
Cores<-makeCluster(detectCores())
```

My PC has an Intel i7-10750H with 6 Cores and 12 total threads.

Let's run the serial loop first..

```{r serial_loopin}
#Baseline time
B4Time<-Sys.time()

#Loop
for (i in 1:4000){
  mean(rnorm(1000000, 10, 3))
}

#Finished loop time
AFTime<-Sys.time()
```

The serial loop took 4 minutes & 13 seconds to run.

How much faster will the parallel loop be? Using a simplified version of Amdahl's Law.. $$
S(n) = \frac{1}{(1-P) + \frac{P}{n}}
$$ .. where $S(n)$ is the theoretical speedup, P is the fraction of the algorithm that can be made parallel, and n is the number of CPU threads..

$$
S(n) = \frac{1}{(1-1) + \frac{1}{12}} = 12
$$ So, theoretically, we should see a 12x increase in speed- or the parallel loop will take \~35 seconds.

Now, let's run the parallel loop..

```{r parallel_loopin, results='hide'}

#Baseline time
Par_B4Time<-Sys.time()

#Loop 
registerDoParallel(Cores) #activate multi-thread

foreach(i = 1:4000, .combine=rbind) %dopar% {
  mean(rnorm(1000000, 10, 3))
}

stopCluster(Cores) #de-activate multi-thread

#Finished loop time
Par_AFTime<-Sys.time()

```

The parallel loop took only 37 seconds, quicker than the serial loop by 3 minutes & 36 seconds! Though, our estimate was quicker by 2 seconds.

My PC runs accessibility-related software that I did not turn off during the parallel loop, which may have been the cause of the 2 second delay compared to the theoretical time calculated.
